---
title: 'Project 2: Health and Economic Impact of Severe Weather'
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```


## Synopsis
The goal of this analysis is to answer two questions in regards to severe storms: 
1. Across the United States, which types of storm events are most harmful with respect to population health?
2. Across the United States, which types of storm events have the greatest economic consequences?

[Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database
will be used to answer the questions. 

## Data Processing


#### Read in Storm Data
The first step is to download the data and read it into a dataframe: 
```{r cache = TRUE}
fileName <- 'repdata_data_StormData.csv.bz2'
if (!file.exists(fileName)){
    download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2')
}
storms <- read.csv(fileName)
head(storms)
```



#### Population Health Analysis
Now that we have the storm data, we will focus on the storms' impact on population
health. Out of the columns in the storms dataframe, FATALITIES and INURIES
impact public health the most. 

In order to find the storms that have the most fatalties and injuries, we will first find the total number of fatalities and injuries per type of storm (EVTYPE). We will also eliminate any storms with 0 fatalities or injuries since these have no measurable affect on public health: 
```{r}
storm_fatalities <- storms %>% 
    group_by(EVTYPE) %>% 
    summarize(total = sum(FATALITIES)) %>%
    filter(total > 0)

storm_injuries <- storms %>% 
    group_by(EVTYPE) %>% 
    summarize(total = sum(INJURIES)) %>%
    filter(total > 0)
```

Next, we will find the outliers in the total number of fatalities and injuries to
find the storms that are most dangerous to public health:

```{r}
fatality_outliers <- boxplot.stats(storm_fatalities$total)$out
top_fatal_storms <- filter(
    storm_fatalities, 
    total %in% fatality_outliers
)
injury_outliers <- boxplot.stats(storm_injuries$total)$out
top_injury_storms <- filter(
    storm_injuries, 
    total %in% injury_outliers
)
```

Note that this approach works because outliers exists only above the median in terms of both fatalities and injuries. If outliers existed in
the lower extrema, then these outliers wouldn't not be representative of storms that have dangerous impact to public health. The following code proves that there are no outliers below the median: 
```{r}
fatality_outliers[fatality_outliers < median(storm_fatalities$total)]
injury_outliers[injury_outliers < median(storm_injuries$total)]

```

Now we will find storms that are most dangerous to public health in terms of both fatalities and injuries by inner joining:
```{r}
pop_health_storms <- inner_join(
    top_fatal_storms, 
    top_injury_storms,
    by = "EVTYPE"
) 
pop_health_storms <-rename(
    pop_health_storms,
    fatalities = total.x, 
    injuries = total.y
)
head(pop_health_storms)
dim(pop_health_storms)
```

Next, we will create a score to quantify negative impact to public health called neg_pop_health_score. Since fatalities are more serious than injuries in terms of public health, they will account for 2/3 the weight of the score, whereas injuries will only account for 1/3 the weight of the score: 
```{r}
pop_health_storms <- mutate(
        pop_health_storms,
        neg_pop_health_score = 2/3 * fatalities + 1/3 * injuries
    ) 
head(pop_health_storms)
dim(pop_health_storms)
```

Next, we will obtain the 5 storms with the highest neg_pop_health_score values:
```{r}
top_5_neg_pop_health_scores <- pop_health_storms %>% 
    select(EVTYPE, neg_pop_health_score) %>% 
    arrange(desc(neg_pop_health_score)) %>% 
    top_n(5)
head(top_5_neg_pop_health_scores)
```

At this point, we have a solid answer of the 5 storms with the worst impact on public health. However, we will do some more processing in order to compare these storms with others storms that have a high number of injuries and/or fatalities. 

We will add a boolean column to the joined data from above called top_5_neg_pop_health to indicate whether a storm is one of the 5 storms with the worst impact on public health: 

```{r}
pop_health_storms <- pop_health_storms %>%
    mutate(top_5_neg_pop_health = ifelse(
        test = EVTYPE %in% top_5_neg_pop_health_scores$EVTYPE,
        TRUE,
        FALSE
    )) %>%
    arrange(desc(neg_pop_health_score))
head(pop_health_storms)
dim(pop_health_storms)
```

Finally, we will melt the joined data so that each storm event will contain a row for fatalitiy count and injury count. This will make plotting with facets easier in the Resutls section: 

```{r}
melted_pop_health_storms <- pivot_longer(
    pop_health_storms, 
    -c(EVTYPE, neg_pop_health_score, top_5_neg_pop_health), 
    names_to = "pop_health_variable", 
    values_to = "total"
)
head(melted_pop_health_storms)
dim(melted_pop_health_storms)
```


### Economy Analysis
We will now run a very similar analysis on the storms' impact on economy. Out of
the columns in the storms dataframe, CROPDMG (crop damage) and PROPDMG (prop damage)
impact the economy the most. 

In order to find the storms that have the most crop damage and property damage, 
we will first find the sum of crop and property damage per type of storm (EVTYPE). 
We will also eliminate any storms with 0 crop and property damage since these have
no measurable affect on the economy: 
```{r}
storm_cropdmg <- storms %>% 
    group_by(EVTYPE) %>% 
    summarize(total = sum(CROPDMG)) %>%
    filter(total > 0)

storm_propdmg <- storms %>% 
    group_by(EVTYPE) %>% 
    summarize(total = sum(PROPDMG)) %>%
    filter(total > 0)
```

Next, we will find the outliers in the total crop and property damage to
find the storms that are most dangerous to the economy:

```{r}
cropdmg_outliers <- boxplot.stats(storm_cropdmg$total)$out
top_cropdmg_storms <- filter(
    storm_cropdmg, 
    total %in% cropdmg_outliers
)
propdmg_outliers <- boxplot.stats(storm_propdmg$total)$out
top_propdmg_storms <- filter(
    storm_propdmg, 
    total %in% propdmg_outliers
)
```

Note that, again, this approach works because outliers exists only above the median in terms of both crop and property damage. If outliers existed in
the lower extrema, then these outliers wouldn't not be representative of storms that have a negative impact on the economy. The following code proves that there are no outliers below the median: 
```{r}
cropdmg_outliers[cropdmg_outliers < median(storm_cropdmg$total)]
propdmg_outliers[propdmg_outliers < median(storm_propdmg$total)]

```

Now we will find storms that are most dangerous to economy in terms of both crop and property damage by inner joining:
```{r}
economy_storms <- inner_join(
    top_cropdmg_storms, 
    top_propdmg_storms,
    by = "EVTYPE"
) 
economy_storms <-rename(
    economy_storms,
    cropdmg = total.x, 
    propdmg = total.y
)
head(economy_storms)
dim(economy_storms)
```

Next, we will create a score to quantify negative impact to the economy called neg_economy_score. Property damage and crop damage both have an equally negative affect on the economy, so each will contribute to 1/2 of the neg_ecomony_score. 
```{r}
economy_storms <- mutate(
        economy_storms,
        neg_economy_score = 1/2 * cropdmg + 1/2 * propdmg
    ) 
head(economy_storms)
dim(economy_storms)
```

Next, we will obtain the 5 storms with the highest neg_economy_score values:
```{r}
top_5_neg_economy_scores <- economy_storms %>% 
    select(EVTYPE, neg_economy_score) %>% 
    arrange(desc(neg_economy_score)) %>% 
    top_n(5)
head(top_5_neg_economy_scores)
```

At this point, we have a solid answer of the 5 storms with the worst impact on the economy. However, we will do some more processing in order to compare these storms with others storms contribute to a large value of property and crop damage. 

We will add a boolean column to the joined data from above called top_5_neg_economy to indicate whether a storm is one of the 5 storms with the worst impact on the economy: 

```{r}
economy_storms <- economy_storms %>%
    mutate(top_5_neg_economy = ifelse(
        test = EVTYPE %in% top_5_neg_economy_scores$EVTYPE,
        TRUE,
        FALSE
    )) %>%
    arrange(desc(neg_economy_score))
head(economy_storms)
dim(economy_storms)
```

Finally, we will melt the joined data so that each storm event will contain a row for property damage and crop damaget. This will make plotting with facets easier in the Resutls section: 

```{r}
melted_economy_storms <- pivot_longer(
    economy_storms, 
    -c(EVTYPE, neg_economy_score, top_5_neg_economy), 
    names_to = "economy_variable", 
    values_to = "total"
)
head(melted_economy_storms)
dim(melted_economy_storms)
```
## Results


#### The Storms with the Worst Impact on Population Health
In the Data Processing section, we found the 5 storms with the worst impact on public health using the neg_pop_health_score value: 
```{r}
head(top_5_neg_pop_health_scores)
```

While these storms have the most negative impact on public health, they don't necessarily have the highest number of fatalities/injuries: 
```{r}
ggplot() + 
    geom_bar(
        aes(EVTYPE, total, fill = top_5_neg_pop_health), 
        data = melted_pop_health_storms, 
        stat = "identity"
    ) + 
    xlab("Event") + 
    theme(axis.text.x = element_text(angle=90)) +     
    facet_grid(
        pop_health_variable ~ ., 
        scales = "free"
    )
```

Looking at the above plot, for example, Flash Floods result in more fatalties than Floods. However, Floods have far more injuries than Flash Floods:

```{r}
flood_data <- filter(pop_health_storms, grepl("FLOOD", EVTYPE))
head(flood_data)
```
This gives Floods a higher neg_pop_health_score, meaning Floods are considered to have worse impact on public health than Flash Floods. 


#### The Storms with the Worst Impact on the Economy
In the Data Processing section, we found the 5 storms with the worst impact on the economy using the neg_economy_score value: 
```{r}
head(top_5_neg_economy_scores)
```

While these storms have the most negative impact on public health, they don't necessarily have the highest amount of crop or property damage: 
```{r}
ggplot() + 
     geom_bar(
         aes(EVTYPE, total, fill = top_5_neg_economy), 
         data = melted_economy_storms, 
         stat = "identity"
     ) + 
     xlab("Event") + 
     theme(axis.text.x = element_text(angle=90)) +     
     facet_grid(
         economy_variable ~ ., 
         scales = "free"
     )
```

Looking at the above plot, for example, Thunderstorm Winds result in more property damage than Hail. However, Hail results in far more crop damage than Thunderstorm Winds:

```{r}
hail_ts_wind_data <- filter(
    economy_storms, 
    EVTYPE %in% c("HAIL", "THUNDERSTORM WINDS")
)
head(hail_ts_wind_data)
```
This gives Hail a higher neg_economy_score, meaning Hail is considered to have worse impact on the economy than Thunderstorm Winds. 

